A pipeline to analyse time-course gene expression data

### 2019-08-09 scripts/manuscript.Rmd
changes to manuscript

- Do we need installation instructions for moanin in this document, since its
  not on CRAN/bioconductor? Also, how its set up now (doing make file, which
  compiles the documentation, etc), probably prohibits use of, e.g.
  `install_github` from `devtools`. If able to call
  `install_github("NelleV/2019timecourse-rnaseq-pipeline",
  subdir="bin/moanin", dependencies=TRUE)` would be simpler (would also
  install the dependencies and get around problems in script)
- I got an error in installation script for  packages:

@@

```{r pca_plots, fig.width=6, fig.height=6}
par(mfrow=c(2, 2))
par(mar=c(4.5, 4.5, 1.5, 1.5))
plot(
    pca_data$x[, c("PC1","PC2")],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
mtext(at=275,text="Colored by Group", line=0.5 ,font=2,adj=0.5)
legend(x=250,y=25,legend=names(ann_colors$Group), fill=ann_colors$Group,bty="n",xpd=NA,yjust=0.5)
par(mar=c(4.5, 5.5, 1.5, .1))
plot(
    pca_data$x[, c("PC3","PC2")],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    ylab="", bty="l", yaxt="n")
axis(2, labels=FALSE)
par(mar=c(4.5, 4.5, 1.5, 1.5))
plot(
    pca_data$x[, c("PC1","PC2")],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
mtext(at=275,text="Colored by Timepoint",line=0.5 ,font=2,adj=0.5)
legend(x=250,y=25,legend=names(ann_colors$Timepoint), fill=ann_colors$Timepoint,bty="n",xpd=NA,yjust=0.5)
par(mar=c(4.5, 5.5, 1.5, .1))
plot(
    pca_data$x[, c("PC3","PC2")],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    ylab="", bty="l", yaxt="n")
axis(2, labels=FALSE)

@@

First, create the contrasts for all timepoints between the two groups of
interest:
```{r make_timepoints_contarts}
contrasts = create_timepoints_contrasts("M", "VL", splines_model)
```

```{r echo=FALSE}
kable(contrasts)
```

Then run the differential expression analysis on all of those timepoints
jointly.

```{r weekly_differential_analysis}
# Define contrast	
weekly_de_analysis = DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)
```

```{r echo=FALSE}
kable(weekly_de_analysis[1:10, ])
```

We can repeat this for each of the conditions

@@

```{r fig.width=6, fig.height=6, echo=FALSE}
par(mfrow=c(2, 2))

time = c(0, 3, 6, 9, 12, 18, 24, 30, 36, 48, 60, 72, 120, 168)

timely_barplot = function(de_analysis, labels, qval_threshold=0.05, title=""){
    qval_colnames = colnames(de_analysis)[
	grepl("qval", colnames(de_analysis))]
    number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)
    names(number_de_genes_per_time) = labels[1]
    barplot(number_de_genes_per_time, col="black", ylim=c(0, 20000),
	    xlab="Timepoint", ylab="Number of DE genes", main=title)

@@

timely_barplot(K_weekly_de_analysis, time, title="Kawasaki")
timely_barplot(C_weekly_de_analysis, time, title="California")
timely_barplot(weekly_de_analysis, time, title="Vietnam low-dose")
timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose")
```

The distribution of number of genes found differentially expressed by
considering each time-point independantly highlights the challenges of such
approach. Timepoint 6H and 18H of the Kawasaki strain have fewer numbers of
genes found differentially expressed than timepoints 3H and 9H: this is likely
due to biological or technical variances for specific genes at specific
timepoints. **EAP: We should add more to this**

As a summary, classic differential expression methods are appropriate for
unordered treatments, but fail to use the temporal structure of the data.

>>>>

- Why `dependicies=FALSE`? (same for the README for the repos). Seems easier to just say dependencies=TRUE and avoid the install.R routine. 
- I got an error in installation script for  packages:

@@

```{r pca_plots, fig.width=6, fig.height=6}
par(mfrow=c(2, 2))
par(mar=c(4.5, 4.5, 1.5, 1.5))
plot(
    pca_data$x[, c("PC1","PC2")],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
mtext(at=275,text="Colored by Group", line=0.5 ,font=2,adj=0.5)
legend(x=250,y=25,legend=names(ann_colors$Group), fill=ann_colors$Group,bty="n",xpd=NA,yjust=0.5)
par(mar=c(4.5, 5.5, 1.5, .1))
plot(
    pca_data$x[, c("PC3","PC2")],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    ylab="", bty="l", yaxt="n")
axis(2, labels=FALSE)
par(mar=c(4.5, 4.5, 1.5, 1.5))
plot(
    pca_data$x[, c("PC1","PC2")],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
mtext(at=275,text="Colored by Timepoint",line=0.5 ,font=2,adj=0.5)
legend(x=250,y=-15,legend=names(ann_colors$Timepoint), fill=ann_colors$Timepoint,bty="n",xpd=NA,yjust=0.5)
par(mar=c(4.5, 5.5, 1.5, .1))
plot(
    pca_data$x[, c("PC3","PC2")],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    ylab="", bty="l", yaxt="n")
axis(2, labels=FALSE)
```

@@

First, create the contrasts for all timepoints between the two groups of
interest:

```{r make_timepoints_contarts}
contrasts = create_timepoints_contrasts("M", "VL", splines_model)
```

This creates a vector of contrasts to be tested, one for each timepoint:

```{r echo=FALSE}
kable(contrasts,col.names="Contrasts")
```
**EAP: Not sure if `kable` is better than just printout?**

Then run the differential expression analysis on all of those timepoints
jointly using the function `DE_timepoints`.

```{r weekly_differential_analysis}
# Define contrast	
VL_weekly_de_analysis = DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)
```
**EAP: Sort of strange to have argument `spines_model` for function that doesn't use splines. Perhaps there is a generic name that could be used for the argument**

This results in a table of the results, where each row corresponds to a gene and the columns correspond to the p-value (`pval`), log-fold change (`lfc`) and adjusted p-value (`qval`); the order of the genes in the table is the same as the input `data`. Here we show the results for the first two timepoints and the first ten genes:

```{r echo=FALSE}
kable(weekly_de_analysis[1:10, 1:6 ],digits=3)
```

**EAP: For the output of this, the column names are in format `M.0-VL.0-pval`, for example. This may be hard to parse algorithmically down the road because you are using the same character for the contrast (`-`) as the separator for the `pval`. So for example, if you wanted to split the contrast name from the `pval` or `lfc` part, it wouldn't be very easy. Maybe underscore `_` would be better separator?**

We will repeat this, comparing each of the treatments to the control ("M") (code not printed here, as it is a replicate of the above, see accompanying Rmarkdown document).

@@

**EAP: Should `timely_barplot` be a function of `moanin`?**

```{r fig.width=6, fig.height=6, echo=FALSE}
timely_barplot = function(de_analysis, labels=NULL, qval_threshold=0.05, title="",...){
    qval_colnames = colnames(de_analysis)[
	grepl("qval", colnames(de_analysis))]
    if(is.null(labels)) labels<-sapply(strsplit(gsub("-qval","",qval_colnames),"\\."),.subset2,3)
    number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)
    barplot(number_de_genes_per_time, col="black", 
            names.arg=labels,ylim=c(0, 20000),
	    xlab="Timepoint", ylab="Number of DE genes", main=title,...)

@@

par(mfrow=c(2, 2))
timely_barplot(K_weekly_de_analysis, time, title="Kawasaki",las=3)
timely_barplot(C_weekly_de_analysis, time, title="California",las=3)
timely_barplot(VL_weekly_de_analysis, time, title="Vietnam low-dose",las=3)
timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose",las=3)
```

Such an analysis can demonstrate some general trends, with clearly more genes being differentially expressed at later time points, and the Vietnam high-dose showing perhaps an earlier onset than the low-dose.

However, the distribution of the number of genes found differentially expressed by
considering each time-point independently highlights the challenges of such
approach. We can see that some timepoints have many less genes found significantly differentially expressed (e.g. timepoint 6H and 18H of the Kawasaki strain). While there may be biological differences at those time points for some genes, it seems unlikely that the large majority of genes differentially expressed at timepoint 3H stop being differentially expressed at 6H and then jump back to being differentially expressed at 9H. A more likely explanation is that there is are some technical or biological artifact about the samples for 6H that is creating higher variation and thus less ability to detect significance.

```{r calculateCombos, echo=FALSE, results="hide"}
#Code for counting the number of unique combinations of time points that are DE
getTimepoint<-function(x){sapply(strsplit(gsub("-qval","",x),"\\."),.subset2,3)}
qval_colnames = colnames(K_weekly_de_analysis)[
grepl("qval", colnames(K_weekly_de_analysis))]
signifCombos<-apply(K_weekly_de_analysis[, qval_colnames],1,function(x){paste(getTimepoint(qval_colnames[which(x<0.05)]),collapse=",")})
signifCombos<-signifCombos[signifCombos!=""]
tabCombos<-table(signifCombos)
```

Another difficulty with such an approach is making sense of the general temporal structure for any particular gene, as different genes will have different combinations of timepoints DE. For the comparison of the Kawasaki strain to the control, for example, there are `length(signifCombos)` genes found DE in some timepoints, and there are `length(tabCombos)` different combinations of timepoints for which they are DE. Some of these make sense, such as DE in timepoints 48H-168H (`tabCombos["48,60,72,120,168"]` genes), but many are very fragmentary. For example there are `tabCombos["48,60,120,168"]` genes which are DE in timepoints 48,60,120,168H, but *not* in the 72H. Many of these genes are likely to have not made the cutoff for significance in 72H, but don't show real differences in the overall trend between 48H-168H.

Here are plots of the first 10 such genes, where we can see that some might show some meaningful changes between 60 and 72H, but others clearly just have a single replicate that is different or increased variability at 72H.

```{r showMissingDEGenes, echo=FALSE}
exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
wh<-which(splines_model$meta$Group %in% c("K","M"))
plot_genes(data[exampleGenes,],splines_model,colors=ann_colors$Group)
```

**EAP: It would be nice to be able to subset in `plot_centroids/plot_genes` to only some of the groups**

As a summary, classic differential expression methods are appropriate for
unordered treatments, but fail to make use of the temporal structure of the data.
